Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-01-11 xrb2/paper.tex
some rephrasing

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. {\color{blue}We saw in \citet{flame_wave1} that 
increasing the 
neutron star rotation rate reduces the horizontal lengthscale of the flame. An estimate of 
this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius may be 
thought of as the scale over which the balance between the Coriolis force and horizontal pressure 
gradient becomes important, and is approximately:
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use the energy generation rate $\dot{e}_\mathrm{nuc}$ at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ for {\tt F500} and {\tt F1000} to discern the horizontal extent of the flame at different rotation rates. Taking the rightmost edge of the bright teal/green region (where significant energy generation is occuring) as the leading edge of the flame in each plot, we see that the horizontal extent of the $\tt{F1000}$ flame appears to be 
reduced compared to the lower rotation $\tt{F500}$ run. From Equation \ref{eqn:rossby}, increasing 
the rotation rate from $500~\mathrm{Hz}$ to $1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and the greater 
confinement from the Coriolis force should reduce the horizontal extent of the flame by a similar 
factor. The Rossby radius is not a perfect measure of this horizontal lengthscale, however, and 
here the difference is not a factor of two. The simulations seem to follow the theoretical
scaling described in Equation \ref{eqn:rossby} more closely at higher rotation rates ($1000~\mathrm{Hz}$ and 
up), based on the results of \citet{flame_wave1}.}

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. {\color{blue} The $\dot{e}_\mathrm{nuc}$ plots in Figure \ref{fig:time_series_enuc_1000} also support this interpretation, with the interior of the {\tt F1000} flame reaching higher $\dot{e}_\mathrm{nuc}$ values than for the {\tt F500} run in Figure \ref{fig:time_series_enuc_500}.} In contrast to {\tt F500} and {\tt F1000}, the 
lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition for {\tt F250} also 
aligns with the reasoning given above, with the lower rotation in this case potentially leading to 
insufficient confinement such that the temperature and density required for ignition is not 
achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) would need to 
take over at lower rotation rates to allow a burst to occur, at least for the initial model used 
here.

@@

\begin{deluxetable}{lrrr}
	\tablecaption{\label{table:flame_speeds} Flame speed and acceleration values measured for each 
	simulation. {\color{blue}$v_0$} is the flame velocity {\color{blue}in the $+r$ direction} at $t 
	= 0$. The initial flame 
	velocities and accelerations are derived from a quadratic least-squares fit to each of the 
	datasets for times $t \gtrsim 20~\mathrm{ms}$. {\color{blue} Using these fit parameters we calculate the velocity at $t = 50~\mathrm{ms}$ ($v_{50}$), at which point the flames have reached steady propagation.}}
	\tablehead{\colhead{run} & \colhead{$v_0$ (cm s$^{-1}$)} & \colhead{$a$ (cm s$^{-2}$)} & \colhead{$v_{50}$ (cm s$^{-1}$)}} 
	\startdata
  {\tt F1000}     & $(3.414 \pm 0.008) \times 10^5$ & $(-1.03 \pm 0.13) \times 10^5$ & $(3.362 \pm 0.010) \times 10^5$\\
  {\tt F500}      & $(3.077 \pm 0.013) \times 10^5$ & $(5.41 \pm 0.19) \times 10^5$ & $(3.347 \pm 0.016) \times 10^5$\\
  {\tt F500\_2E8} & $(3.760 \pm 0.015) \times 10^5$ & $(1.74 \pm 0.25) \times 10^5$ & $(3.847 \pm 0.020) \times 10^5$\\
  {\tt F500\_3E8} & $(-5.293 \pm 0.100) \times 10^5$ & $(3.57 \pm 0.02) \times 10^7$ & $(1.258 \pm 0.015) \times 10^6$\\
	\enddata
	%
\end{deluxetable}

@@

Figure~\ref{fig:flame_speeds_1} gives position vs.\ time for the $\tt{F500}$ and $\tt{F1000}$ runs 
(blue and orange, respectively) to show the dependence on rotation rate. {\color{blue}In 
\citet{flame_wave1}, we 
used linear least-squares fits to our position vs.\ time data to estimate the propagation velocity. 
For this study we instead fit a quadratic curve to each set of data, as some of the hotter $500~\mathrm{Hz}$ 
runs exhibit significant acceleration. We discard the data points with $t \lesssim 20~\mathrm{ms}$, 
since these correspond to the initial transient period before the flame has begun to propagate 
steadily.} The fit parameters yield estimates of the flame acceleration and velocity at $t = 0$, 
which are provided in Table~\ref{table:flame_speeds}. {\color{blue}Note that the listed velocity is 
only 
a parameter that may be used to calculate the flame speed at an arbitrary time. It is not an 
estimate of the true initial velocity of the flame, since the flame has not achieved ignition yet 
at $t = 0$. We use the fit parameters to calculate the flame speeds at $t = 50~\mathrm{ms}$ (when the flame has reached steady propagation), given in the fourth column of Table \ref{table:flame_speeds}.} \MarginPar{KE: Outside of updating the data in the table, is there anything else we 
want to add to this para?}

@@

\MarginPar{NF: should I include a plot for this or is it ok to just say it?}
Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the characteristic flame structure shown in $\bar{A}$ plots for the lower temperature $500~\mathrm{Hz}$ runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. {\color{blue}A similar effect is also visible (to a lesser extent) in the $\tt{F500\_3E8}$ plot in Figure \ref{fig:compare_500Hz_abar}, with a faint haze of material rising above the flame. This effect, however, may be numerical rather than physical, resulting from the boundary conditions imposed at the axis of symmetry for these simulations.} $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}). This is not surprising, since as we have shown in our simulations with differing initial temperature structures, the burning is very temperature-sensitive.

>>>>

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. {\color{blue}We saw in \citet{flame_wave1} that 
increasing the 
neutron star rotation rate reduces the horizontal lengthscale of the flame. An estimate of 
this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius may be 
thought of as the scale over which the balance between the Coriolis force and horizontal pressure 
gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use the energy generation rate $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region (where significant energy generation is occuring) as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). From Equation \ref{eqn:rossby}, we can see that increasing 
the rotation rate from $500~\mathrm{Hz}$ to $1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater 
confinement from the Coriolis force should reduce the horizontal extent of the flame by a similar 
factor. However, the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our simulations we see that this scaling does not work so well for all rotation rates. The simulations seem to follow the theoretical
scaling described in Equation \ref{eqn:rossby} more closely at higher rotation rates ($1000~\mathrm{Hz}$ and 
higher), based on the results of \citet{flame_wave1}.}

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. {\color{blue} The $\dot{e}_\mathrm{nuc}$ plots in Figure \ref{fig:time_series_enuc_1000} also support this interpretation, with the interior\MarginPar{What is meant by interior here?? The edge? The part nearest the crust?} of the {\tt F1000} flame reaching higher $\dot{e}_\mathrm{nuc}$ values than for the {\tt F500} run in Figure \ref{fig:time_series_enuc_500}.} In contrast to {\tt F500} and {\tt F1000}, the 
lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition for {\tt F250} also 
aligns with the reasoning given above, with the lower rotation in this case potentially leading to 
insufficient confinement such that the temperature and density required for ignition is not 
achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) would need to 
take over at lower rotation rates to allow a burst to occur, at least for the initial model used 
here.

@@

\begin{deluxetable}{lrrr}
	\tablecaption{\label{table:flame_speeds} Flame speed and acceleration values measured for each 
	simulation. {\color{blue}$v_0$} is the flame velocity {\color{blue}in the $+r$ direction} at $t 
	= 0~\mathrm{ms}$. The initial flame 
	velocities and accelerations are derived from a quadratic least-squares fit to each of the 
	datasets for times $t \gtrsim 20~\mathrm{ms}$. {\color{blue} Using these fit parameters we calculate the velocity at $t = 50~\mathrm{ms}$, $v_{50}$, at which point the flames have reached steady propagation.}}
	\tablehead{\colhead{run} & \colhead{$v_0$ (cm s$^{-1}$)} & \colhead{$a$ (cm s$^{-2}$)} & \colhead{$v_{50}$ (cm s$^{-1}$)}} 
	\startdata
  {\tt F1000}     & $(3.414 \pm 0.008) \times 10^5$ & $(-1.03 \pm 0.13) \times 10^5$ & $(3.362 \pm 0.010) \times 10^5$\\
  {\tt F500}      & $(3.077 \pm 0.013) \times 10^5$ & $(5.41 \pm 0.19) \times 10^5$ & $(3.347 \pm 0.016) \times 10^5$\\
  {\tt F500\_2E8} & $(3.760 \pm 0.015) \times 10^5$ & $(1.74 \pm 0.25) \times 10^5$ & $(3.847 \pm 0.020) \times 10^5$\\
  {\tt F500\_3E8} & $(-5.293 \pm 0.100) \times 10^5$ & $(3.57 \pm 0.02) \times 10^7$ & $(1.258 \pm 0.015) \times 10^6$\\
	\enddata
	%
\end{deluxetable}

@@

Figure~\ref{fig:flame_speeds_1} gives the position of the flame front as a function of time for the $\tt{F500}$ and $\tt{F1000}$ runs 
(blue and orange, respectively) to show the dependence on rotation rate. {\color{blue}In 
\citet{flame_wave1}, we 
applied a linear least-squares fit to the flame front position as a function of time to estimate the propagation velocity. 
For this study we instead fit a quadratic curve to each set of data, as some of the hotter $500~\mathrm{Hz}$ 
runs exhibit significant acceleration. We discarded the data points with $t \lesssim 20~\mathrm{ms}$, 
since these correspond to the initial transient period before the flame has begun to propagate 
steadily.} The fit parameters yield estimates of the flame acceleration and velocity at $t = 0~\mathrm{ms}$, 
which are provided in Table~\ref{table:flame_speeds}. {\color{blue}Note that the listed velocity is 
only 
a parameter that may be used to calculate the flame speed at an arbitrary time. It is not an 
estimate of the true initial velocity of the flame, since the flame has not achieved ignition yet 
at $t = 0~\mathrm{ms}$. We use the fit parameters to calculate the flame speeds at $t = 50~\mathrm{ms}$ (when the flame has reached steady propagation), given in the fourth column of Table \ref{table:flame_speeds}.} \MarginPar{KE: Outside of updating the data in the table, is there anything else we 
want to add to this para?}

@@

\MarginPar{NF: should I include a plot for this or is it ok to just say it?}
Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the characteristic flame structure shown in $\bar{A}$ plots for the lower temperature $500~\mathrm{Hz}$ runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. {\color{blue}A similar effect is also visible (to a lesser extent) in the $\tt{F500\_3E8}$ plot in Figure \ref{fig:compare_500Hz_abar}, with a faint haze of material rising above the flame. However, this is likely to be a numerical artefact rather than a true physical effect, and a result of the boundary conditions imposed at the axis of symmetry for these simulations.} $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}). This is not surprising, since as we have shown in our simulations with differing initial temperature structures, the burning is very temperature-sensitive.
